\subsection {PTRACERS Package} 
\label{sec:pkg:ptracers}
\begin{rawhtml}
<!-- CMIREDIR:package_ptracers: -->
\end{rawhtml}

\subsubsection {Introduction}

This is a "passive" tracer package. Passive here means that the
tracers don't affect the density of the water (as opposed to
temperature and salinity) so no not actively affect the 
physics of the ocean. Tracers are initialized, advected, diffused
and various outputs are taken care of in this package.
For methods to add additional sources and sinks of tracers use the pkg/gchem
(section \ref{sec:pkg:gchem}).

Can use up tp 3843 tracers. But can not use pkg/diagnostics with more 
than about 90 tracers.
Use utils/matlab/ioLb2num.m and num2ioLb.m to find correspondence
between tracer number and tracer designation in the code for more
than 99 tracers (since
tracers only have two digit designations).

\subsubsection {Equations}

\subsubsection {Key subroutines and parameters}

The only code you shoul dhave to modify is:
{\bf PTRACERS\_SIZE.h} where you need to set in the number of tracers to
be used in the experiment:  PTRACERS\_num.

\vspace{.5cm}


\noindent
{bf RUN TIME PARAMETERS  SET IN {\bf data.ptracers}}: \\
$\bullet$ {\bf PTRACERS\_Iter0} 
which is the integer timestep when the tracer experiment
 is initialized. If  nIter0 $=$  PTRACERS\_Iter0 then the tracers
 are initialized to zero or from initial files. If nIter0 $>$
  PTRACERS\_Iter0 then tracers (and previous timestep tendency terms)
  are read in from a the ptracers pickup file. Note that tracers
  of zeros will be carried around if nIter0 $<$  PTRACERS\_Iter0.\\
$\bullet$ {\bf PTRACERS\_numInUse}: number of tracers to be used in the
run (needs to be $<=$ PTRACERS\_num set in PTRACERS\_SIZE.h)\\
$\bullet$  {\bf PTRACERS\_dumpFreq}: defaults to dumpFreq (set in data)\\
$\bullet$  {\bf PTRACERS\_taveFreq}: defaults to taveFreq (set in data)\\
$\bullet$  {\bf PTRACERS\_monitorFreq}: defaults to monitorFreq (set in data)\\
$\bullet$  {\bf PTRACERS\_timeave\_mnc}: needs useMNC, timeave\_mnc, default to false\\
$\bullet$  {\bf PTRACERS\_snapshot\_mnc}:  needs useMNC, snapshot\_mnc, default to false \\
$\bullet$  {\bf PTRACERS\_monitor\_mnc}: needs useMNC, monitor\_mnc, default to false\\
$\bullet$  {\bf PTRACERS\_pickup\_write\_mnc}:  needs useMNC, pickup\_write\_mnc, default to false\\
$\bullet$  {\bf PTRACERS\_pickup\_read\_mnc}: needs useMNC, pickup\_read\_mnc, default to false\\
$\bullet$  {\bf PTRACERS\_useRecords}: defaults to false. If true, will write 
all tracers in a single file, otherwise each tracer in a seperate file.\\

\vspace{.5cm}

The following can be set for each tracer (tracer number iTrc):\\
$\bullet$  {\bf PTRACERS\_advScheme(iTrc)} will default to saltAdvScheme (set in data). For other options see Table \ref {tab:advectionShemes_summary}.\\
$\bullet$  {\bf PTRACERS\_ImplVertAdv(iTrc)}: implicit vertical advection flag,
 default to .FALSE.\\
$\bullet$  {\bf PTRACERS\_diffKh(iTrc)}: horizontal Laplacian Diffusivity,
dafaults to diffKhS (set in data).\\
$\bullet$  {\bf PTRACERS\_diffK4(iTrc)}: Biharmonic Diffusivity, defaults to
diffK4S (set in data).\\
$\bullet$  {\bf PTRACERS\_diffKr(iTrc)}: vertical diffusion, defaults to un-set.\\
$\bullet$  {\bf PTRACERS\_diffKrNr(k,iTrc)}: level specific vertical diffusion,
defaults to diffKrNrS. Will be set to PTRACERS\_diffKr if this is set.\\
$\bullet$  {\bf PTRACERS\_ref(k,iTrc)}: reference tracer value for each level k, 
defaults to 0. Currently only used for
dilution/concentration of tracers at surface if  PTRACERS\_EvPrRn(iTrc) is
set and convertFW2Salt (set in data) is set to something other than -1
(note default is convertFW2Salt=35).\\
$\bullet$  {\bf PTRACERS\_EvPrRn(iTrc)}: tracer concentration in freshwater.
Needed for calculation of dilution/concentration in surface layer due to
freshwater addition/evaporation. Defaults to un-set in which case no
dilution/concentration occurs.\\
$\bullet$  {\bf PTRACERS\_useGMRedi(iTrc)}: apply GM or not. Defaults to
useGMREdi.\\
$\bullet$  {\bf PTRACERS\_useKPP(iTrc)}: apply KPP or not. Defaults to
useKPP.\\
$\bullet$  {\bf PTRACERS\_initialFile(iTrc)}: file with initial tracer 
concentration. Will be used if PTRACERS\_Iter0 $=$ nIter0. Default
is no name, in which case tracer is initialised as zero. If 
 PTRACERS\_Iter0 $<$ nIter0, then tracer concentration will come
from pickup\_ptracer.\\
$\bullet$  {\bf PTRACERS\_names(iTrc)}: tracer name. Needed for netcdf.
Defaults to nothing.\\
$\bullet$  {\bf PTRACERS\_long\_names(iTrc)}: optional name in long form
of tracer.\\
$\bullet$  {\bf PTRACERS\_units(iTrc)}: optional units of tracer.
\\


\subsubsection{PTRACERS Diagnostics}
\label{sec:pkg:ptracers:diagnostics}

Note that these will only work for 90 or less tracers (some
problems with the numbering/designation over this number)

{\footnotesize
\begin{verbatim}

------------------------------------------------------------------------
<-Name->|Levs|<-parsing code->|<--  Units   -->|<- Tile (max=80c) 
------------------------------------------------------------------------
TRAC01  | 15 |SM P    MR      |mol C/m         |Mass-Weighted Dissolved Inorganic Carbon
UTRAC01 | 15 |UU   171MR      |mol C/m.m/s     |Zonal Mass-Weighted Transp of Dissolved Inorganic Carbon
VTRAC01 | 15 |VV   170MR      |mol C/m.m/s     |Merid Mass-Weighted Transp of Dissolved Inorganic Carbon
WTRAC01 | 15 |WM      MR      |mol C/m.m/s     |Vert  Mass-Weighted Transp of Dissolved Inorganic Carbon
ADVrTr01| 15 |WM      LR      |mol C/m.m^3/s   |Vertical   Advective Flux of Dissolved Inorganic Carbon
ADVxTr01| 15 |UU   175MR      |mol C/m.m^3/s   |Zonal      Advective Flux of Dissolved Inorganic Carbon
ADVyTr01| 15 |VV   174MR      |mol C/m.m^3/s   |Meridional Advective Flux of Dissolved Inorganic Carbon
DFrETr01| 15 |WM      LR      |mol C/m.m^3/s   |Vertical Diffusive Flux of Dissolved Inorganic Carbon (Explicit part)
DIFxTr01| 15 |UU   178MR      |mol C/m.m^3/s   |Zonal      Diffusive Flux of Dissolved Inorganic Carbon
DIFyTr01| 15 |VV   177MR      |mol C/m.m^3/s   |Meridional Diffusive Flux of Dissolved Inorganic Carbon
DFrITr01| 15 |WM      LR      |mol C/m.m^3/s   |Vertical Diffusive Flux of Dissolved Inorganic Carbon (Implicit part)
TRAC02  | 15 |SM P    MR      |mol eq/         |Mass-Weighted Alkalinity
UTRAC02 | 15 |UU   182MR      |mol eq/.m/s     |Zonal Mass-Weighted Transp of Alkalinity
VTRAC02 | 15 |VV   181MR      |mol eq/.m/s     |Merid Mass-Weighted Transp of Alkalinity
WTRAC02 | 15 |WM      MR      |mol eq/.m/s     |Vert  Mass-Weighted Transp of Alkalinity
ADVrTr02| 15 |WM      LR      |mol eq/.m^3/s   |Vertical   Advective Flux of Alkalinity
ADVxTr02| 15 |UU   186MR      |mol eq/.m^3/s   |Zonal      Advective Flux of Alkalinity
ADVyTr02| 15 |VV   185MR      |mol eq/.m^3/s   |Meridional Advective Flux of Alkalinity
DFrETr02| 15 |WM      LR      |mol eq/.m^3/s   |Vertical Diffusive Flux of Alkalinity (Explicit part)
DIFxTr02| 15 |UU   189MR      |mol eq/.m^3/s   |Zonal      Diffusive Flux of Alkalinity
DIFyTr02| 15 |VV   188MR      |mol eq/.m^3/s   |Meridional Diffusive Flux of Alkalinity
DFrITr02| 15 |WM      LR      |mol eq/.m^3/s   |Vertical Diffusive Flux of Alkalinity (Implicit part)
TRAC03  | 15 |SM P    MR      |mol P/m         |Mass-Weighted Phosphate
UTRAC03 | 15 |UU   193MR      |mol P/m.m/s     |Zonal Mass-Weighted Transp of Phosphate
VTRAC03 | 15 |VV   192MR      |mol P/m.m/s     |Merid Mass-Weighted Transp of Phosphate
WTRAC03 | 15 |WM      MR      |mol P/m.m/s     |Vert  Mass-Weighted Transp of Phosphate
ADVrTr03| 15 |WM      LR      |mol P/m.m^3/s   |Vertical   Advective Flux of Phosphate
ADVxTr03| 15 |UU   197MR      |mol P/m.m^3/s   |Zonal      Advective Flux of Phosphate
ADVyTr03| 15 |VV   196MR      |mol P/m.m^3/s   |Meridional Advective Flux of Phosphate
DFrETr03| 15 |WM      LR      |mol P/m.m^3/s   |Vertical Diffusive Flux of Phosphate (Explicit part)
DIFxTr03| 15 |UU   200MR      |mol P/m.m^3/s   |Zonal      Diffusive Flux of Phosphate
------------------------------------------------------------------------
<-Name->|Levs|<-parsing code->|<--  Units   -->|<- Tile (max=80c) 
------------------------------------------------------------------------
DIFyTr03| 15 |VV   199MR      |mol P/m.m^3/s   |Meridional Diffusive Flux of Phosphate
DFrITr03| 15 |WM      LR      |mol P/m.m^3/s   |Vertical Diffusive Flux of Phosphate (Implicit part)
TRAC04  | 15 |SM P    MR      |mol P/m         |Mass-Weighted Dissolved Organic Phosphorus
UTRAC04 | 15 |UU   204MR      |mol P/m.m/s     |Zonal Mass-Weighted Transp of Dissolved Organic Phosphorus
VTRAC04 | 15 |VV   203MR      |mol P/m.m/s     |Merid Mass-Weighted Transp of Dissolved Organic Phosphorus
WTRAC04 | 15 |WM      MR      |mol P/m.m/s     |Vert  Mass-Weighted Transp of Dissolved Organic Phosphorus
ADVrTr04| 15 |WM      LR      |mol P/m.m^3/s   |Vertical   Advective Flux of Dissolved Organic Phosphorus
ADVxTr04| 15 |UU   208MR      |mol P/m.m^3/s   |Zonal      Advective Flux of Dissolved Organic Phosphorus
ADVyTr04| 15 |VV   207MR      |mol P/m.m^3/s   |Meridional Advective Flux of Dissolved Organic Phosphorus
DFrETr04| 15 |WM      LR      |mol P/m.m^3/s   |Vertical Diffusive Flux of Dissolved Organic Phosphorus (Explicit part)
DIFxTr04| 15 |UU   211MR      |mol P/m.m^3/s   |Zonal      Diffusive Flux of Dissolved Organic Phosphorus
DIFyTr04| 15 |VV   210MR      |mol P/m.m^3/s   |Meridional Diffusive Flux of Dissolved Organic Phosphorus
DFrITr04| 15 |WM      LR      |mol P/m.m^3/s   |Vertical Diffusive Flux of Dissolved Organic Phosphorus (Implicit part)
TRAC05  | 15 |SM P    MR      |mol O/m         |Mass-Weighted Dissolved Oxygen
UTRAC05 | 15 |UU   215MR      |mol O/m.m/s     |Zonal Mass-Weighted Transp of Dissolved Oxygen
VTRAC05 | 15 |VV   214MR      |mol O/m.m/s     |Merid Mass-Weighted Transp of Dissolved Oxygen
WTRAC05 | 15 |WM      MR      |mol O/m.m/s     |Vert  Mass-Weighted Transp of Dissolved Oxygen
ADVrTr05| 15 |WM      LR      |mol O/m.m^3/s   |Vertical   Advective Flux of Dissolved Oxygen
ADVxTr05| 15 |UU   219MR      |mol O/m.m^3/s   |Zonal      Advective Flux of Dissolved Oxygen
ADVyTr05| 15 |VV   218MR      |mol O/m.m^3/s   |Meridional Advective Flux of Dissolved Oxygen
DFrETr05| 15 |WM      LR      |mol O/m.m^3/s   |Vertical Diffusive Flux of Dissolved Oxygen (Explicit part)
DIFxTr05| 15 |UU   222MR      |mol O/m.m^3/s   |Zonal      Diffusive Flux of Dissolved Oxygen
DIFyTr05| 15 |VV   221MR      |mol O/m.m^3/s   |Meridional Diffusive Flux of Dissolved Oxygen
DFrITr05| 15 |WM      LR      |mol O/m.m^3/s   |Vertical Diffusive Flux of Dissolved Oxygen (Implicit part)

\end{verbatim}
}

\subsubsection{Do's and Don'ts}

\subsubsection{Reference Material}

