\section{Pre-- and Post--Processing Scripts and Utilities}

There are numerous tools for pre-processing data, converting model
output and analysis written in Matlab, fortran (f77 and f90) and perl.
As yet they remain undocumented although many are self-documenting
(Matlab routines have "help" written into them).

Here we'll summarize what is available but this is an ever growing
resource so this may not cover everything that is out there:

\subsection{Utilities supplied with the model}

We supply some basic scripts with the model to facilitate conversion
or reading of data into analysis software.

\subsubsection{utils/scripts}

In the directory {\em utils/scripts} you will find {\em joinds} and
{\em joinmds}: these are perl scripts used from joining the multi-part
files created by MITgcm. {\bf Use {\em joinmds} always}. You will only
need {\em joinds} if you are working with output older than two years
(prior to c23).

\subsubsection{utils/matlab}

In the directory {\em utils/matlab} you will find several Matlab
scripts (.m or dot-em files). The priniciple script is {\em rdmds.m}
used for reading the multi-part model output files in to matlab. Place
the scripts in your matlab path or change the path appropriately, then
at the matlab prompt type:
\begin{verbatim}
  >> help rdmds
\end{verbatim}
to get help on how to use rdmds.

Another useful script scans the terminal output file for "monitor"
information.

Most other scripts are for working in the curvilinear coordinate
systems which as yet are unpublished and undocumented.

\subsubsection{MNC Utils}

The MITgcm netCDF integration is relatively new and the tools used to
work on netCDF data sets are developing.  The following scripts and
utilities have been written to help manipulate MNC (netCDF) files:
\begin{description}
\item[Tile Assembly:] a matlab script {\tt
    utils/matlab/mnc\_assembly.m} is available for spatially
  ``assembling'' MNC output.  A convenience wrapper script called {\tt
    utils/matlab/gluemnc.m} is also provided.  Please use the matlab
  help facility for more information.

\item[gmt:] As MITgcm evolves to handle more complicated domains and
  topologies, a suite of matlab tools is being written to more
  gracefully handle the model files.  This suite is called ``gmt''
  which refers to ``generalized model topology'' pre-/post-processing.
  Currently, this directory contains a matlab script {\tt
    utils/matlab/gmt/rdnctiles.m} that is able to read the MNC-created
  netCDF files for any domain.  Additional scripts are being created
  that will work with these fields on a per-tile basis.

\end{description}


\subsection{Pre-processing software}

There is a suite of pre-processing software for intepolating
bathymetry and forcing data, written by Adcroft and Biastoch. At some
point, these will be made available for download. If you are in need
of such software, contact one of them.

